---
title: "Bike_Sharing_Case_Study"
author: "Carmelo Pannetto"
date: "31/10/2021"
output: html_document
---


# Install  required packages.
---

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
```
# Load required packages.
---

```{r}
library(tidyverse)  
library(lubridate)  
library(ggplot2)  
library(dplyr) 
```

#=====================
# STEP 1: COLLECT DATA
#=====================
Data has been taken from <https://divvy-tripdata.s3.amazonaws.com/index.html> under Motivate International Inc. under this license
[link](https://www.divvybikes.com/data-license-agreement)

Set up the directory where to upload from, and save the data .( changing slash from \ to /)

```{r}
getwd() #tell us the directory we're working in.
setwd("C:/Users/Melo/Documents/Capstone/Dataset/Trips/Last_year_trips")
```

Upload data:
```{r}
trip_2020_10<- read.csv("Trip_2020_10.csv")
trip_2020_11<- read.csv("Trip_2020_11.csv")
trip_2020_12<- read.csv("Trip_2020_12.csv")
trip_2021_01<- read.csv("Trip_2021_01.csv")
trip_2021_02<- read.csv("Trip_2021_02.csv")
trip_2021_03<- read.csv("Trip_2021_03.csv")
trip_2021_04<- read.csv("Trip_2021_04.csv")
trip_2021_05<- read.csv("Trip_2021_05.csv")
trip_2021_06<- read.csv("Trip_2021_06.csv")
trip_2021_07<- read.csv("Trip_2021_07.csv")
trip_2021_08<- read.csv("Trip_2021_08.csv")
trip_2021_09<- read.csv("Trip_2021_09.csv")
```



#=====================
# STEP 2: WRANGLING DATA AND COMBINE INTO A SINGLE FILE
#=====================

Now Checking for consistance before creating an unique dataframe.
```{r}
str(trip_2020_10)  
str(trip_2020_11)
str(trip_2020_12)
str(trip_2021_01)
str(trip_2021_02)
str(trip_2021_03)
str(trip_2021_04)
str(trip_2021_05)
str(trip_2021_06)
str(trip_2021_07)
str(trip_2021_08)
str(trip_2021_09)
```


I've found that 2 variable from 2 differents dataset are in the wrong format for our dataframe, so i am changing them.
```{r}
#make the variable consistent
trip_2020_10<- mutate(trip_2020_10,start_station_id = as.character(start_station_id)) 
trip_2020_10<- mutate(trip_2020_10,end_station_id = as.character(end_station_id))
trip_2020_11<- mutate(trip_2020_11,start_station_id = as.character(start_station_id))
trip_2020_11<- mutate(trip_2020_11,end_station_id = as.character(end_station_id))

```

Creating a new dataframe that include all our trips from different months:
```{r}
all_trips<- bind_rows(trip_2020_10,trip_2020_11,trip_2020_12,trip_2021_01,  
                      trip_2021_02,trip_2021_03,trip_2021_04,trip_2021_05,
                      trip_2021_06,trip_2021_07,trip_2021_08,trip_2021_09)
```

Renaming column so it's easier to understand what we're working with:
```{r}
all_trips<- rename(all_trips,bike_type = rideable_type, 
                   user_type = member_casual)
```

Saving our new database:
```{r}
save(all_trips, file = "all_trips")
```

 
#=====================
# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#=====================

Inspect the new dataframe:
```{r}
colnames(all_trips) #Give us the columns names
nrow(all_trips) #How many rows
dim(all_trips) #Dimension of dataframe
head(all_trips)#Show the first 6 rows of the dataframe
tail(all_trips)#Show the last 6 rows of the dataframe
str(all_trips) #See list of columns and data types (numeric, character, etc)
summary(all_trips) #Statistical summary of data.
```

I've created new columns to work with data taking time period reference ( day, months year, day of week) :

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Format data to work with function:
```{r}
all_trips$ended_at <- as_datetime(as.character(all_trips$ended_at))#formatting as date for using difftime function
all_trips$started_at <- as_datetime(as.character(all_trips$started_at))
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length)) # now we create the ride length, using the diff function (end time minuts start time ). the ride_length is formatted a char, so well go to convert it as a number
is.numeric(all_trips$ride_length)
```

Removing useless data, test trips, length less the 0 seconds and NA values:
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "WATSON TESTING - DIVVY" | all_trips$ride_length<0 ),] 
all_trips_v2<- na.omit(all_trips_v2) 
```

Save new clean dataframe:
```{r}
save(all_trips_v2, file =" all_trips_v2")
```

#=====================
# STEP 4: CONDUCT ANALYSIS ON THE NEW CLEAN DATAFRAME
#=====================

Now we can conduct analysis on our dataframe:

```{r}
all_trips_v2 %>%  #check how many causal user and how many member we have.
    group_by(user_type)%>%
    count(user_type == "casual")
```

```{r}
summary(all_trips_v2$ride_length) #calculate max, min, mean and median
```

Comparing users type bike activity:
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type, FUN = mean) 
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type, FUN = min)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type, FUN = median)
```

The average ride time by day for each user:
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type + all_trips_v2$day_of_week, FUN = mean)#average ride by 2 different user by weekday
```
Fixing the day's order:
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, #ordering weekdays
                                    levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

Doing the same operation as before but now with days fixed:

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type + all_trips_v2$day_of_week, FUN = mean)
```

Analyze ridership data by type and weekday:
```{r}
all_trips_v2 %>%     
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(user_type, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(user_type, weekday)								# sorts
```

Visualize the number of rides by User type by weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(user_type, weekday) %>% 
  summarize(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(user_type, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = user_type)) +
  geom_col(position = "dodge")+ 
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Number of Rides by Days and User Type",subtitle = "Members versus Casual ") +
  ylab("Number of Rides") +
  xlab("Day of Week")
```

Visualize the number of rides by User type by month
```{r}
all_trips_v2 %>% 
  mutate(month = month(started_at, label = TRUE)) %>% 
  group_by(user_type, month) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(user_type, month) %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = user_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Number of Rides by Month and User Type",
       subtitle = "Members versus Casual") +
  ylab("Number of Rides") +
  xlab("Months")
```

Create a visualization for average duration by weekday :
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(user_type, weekday) %>% 
  summarize(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(user_type, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = user_type)) +
  geom_col(position = "dodge")+
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Average Duration of Rides by Days and User Type",subtitle = "Members versus Casual") +
  ylab("Average Duration of Rides (min)") +
  xlab("Day of Week")
```

Create a visualization for average duration by month :
```{r}
all_trips_v2 %>% 
  mutate(month= month(started_at, label = TRUE)) %>% 
  group_by(user_type, month) %>% 
  summarize(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(user_type, month)  %>% 
  ggplot(aes(x = month, y = average_duration, fill = user_type)) +
  geom_col(position = "dodge")+
  scale_fill_manual(values = c("#CC6633","#6699CC")) +
  labs(title = "Average Duration of Rides by Month and User Type",subtitle = "Members versus Casual") +
  ylab("Average Duration of Rides (min)") +
  xlab("Months")
```



#=====================
# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
#=====================

Create a .csv file to export and work further with our dataset:
```{r}
results <- aggregate( all_trips_v2$ride_length ~ all_trips_v2$user_type + all_trips_v2$day_of_week, FUN = mean)#creating a var to export our results and save the result as a .csv file
write.csv(results, file = "C:/Users/Melo/Documents/Capstone/Dataset/Trips/summary_rides.csv")
```

```{r}
rides_month <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$user_type +
                           all_trips_v2$month,FUN = mean)
write.csv(rides_month, file = 'C:/Users/Melo/Documents/Capstone/Dataset/Trips/avg_length_by_month.csv')

```
Create a slim .csv to upload it on tableau

```{r}
tableau_trips<- all_trips_v2%>%
  select(bike_type,user_type,date,year,month,day,day_of_week,ride_length)
write.csv(tableau_trips, file = 'C:/Users/Melo/Documents/Capstone/Dataset/Trips/tableau_trips.csv')
```

